---
title: "Class 18"
author: "Howard Wang"
date: "11/27/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Install packages
```{r}
# BiocManager::install( c("GenomicDataCommons", "TCGAbiolinks", "maftools", "msa") )
```

## Once installed, load the packages, as usual.
```{r}
library(GenomicDataCommons)
library(TCGAbiolinks)
library(maftools)
```

## Now lets check on GDC status:
```{r}
status()
```

## Querying the GDC from R
```{r}
projects <- getGDCprojects()
head(projects)
```

## Part 2 Identifing sites of mutation

```{r}
library(bio3d)
seqs <- read.fasta("~/Downloads/lecture18_sequences.fa")
seqs
```

## Calculate positional identity scores
```{r}
ide <- conserv(seqs$ali, method="identity")
ide
mutant.sites <- which(ide < 1) 
```
## Exclude gap possitions from analysis

```{r}
gaps <- gap.inspect(seqs)
mutant.sites <- mutant.sites[mutant.sites %in% gaps$f.inds]
mutant.sites
```

## Make a "names" label for our output sequences (one per mutant)

```{r}
mutant.names <- paste0(seqs$ali["P53_wt",mutant.sites],
                       mutant.sites,
                       seqs$ali["P53_mutant",mutant.sites])

mutant.names
```

## Sequence positions surounding each mutant site
```{r}
start.position <- mutant.sites - 8
end.position <-  mutant.sites + 8
```

# Blank matrix to store sub-sequences
```{r}
store.seqs <- matrix("-", nrow=length(mutant.sites), ncol=17)
rownames(store.seqs) <- mutant.names
```

## Extract each sub-sequence
```{r}
for(i in 1:length(mutant.sites)) {
  store.seqs[i,] <- seqs$ali["P53_mutant",start.position[i]:end.position[i]]
}
store.seqs
```

## First blank out the gap positions 
```{r}
store.seqs[store.seqs == "-"] <- ""
```

## Output a FASTA file for further analysis
```{r}
write.fasta(seqs=store.seqs, ids=mutant.names, file="subsequences.fa")
```
































































































































































